Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-08-13 papers/wdmerger_I/paper.tex
fix en dashes -> em dashes

The first attempts to model the results of the merger process came in the
1980s. \cite{nomotoiben:1985} demonstrated that off-center carbon
ignition would occur in the more massive white dwarf as it accreted
mass near the Eddington rate from the less massive white dwarf
overflowing its Roche lobe. \cite{saionomoto:1985} tracked the
evolution of the flame and found that it propagated quiescently into
the center, converting the carbon-oxygen white dwarf into an
oxygen-neon-magnesium white dwarf. This would then be followed by
collapse into a neutron star -- a result with significantly different
observational properties compared to a SN Ia. This scenario, termed
accretion-induced collapse, would be avoided only if the accretion
rate were well below the Eddington rate. \cite{tutukov_yungelson:1979}
observed that this could happen if the mass loss from the secondary
was higher than the Eddington rate and thus the accreted material
formed an accretion disc, which might rain down on the primary more
slowly. The main finding was that double degenerate systems would not
obviously lead to Type Ia supernovae.

@@

A recent shift in perspective on this problem started around 2010.
\cite{pakmor:2010} used the SPH method to study the merger of 
equal-mass ($0.9\ \msolar$) carbon-oxygen white dwarfs and found 
that a hotspot was generated near the surface of the primary 
white dwarf. They argued that this region had a temperature 
and density sufficient to trigger a thermonuclear
detonation. They inserted a detonation which propagated throughout 
the system. They found that the result would observationally 
appear as a subluminous Type Ia supernova. This was the first time 
a DD simulation successfully reproduced at least some characteristics of a SN
Ia. \cite{pakmor:2011} tried a few different mass combinations and
found empirically that this would hold as long as the secondary was at
least 80\% as massive as the primary. These events, where the merger
process resulted in the detonation of the system during the merger
coalescence -- avoiding the much longer time-scale evolution -- were
termed ``violent'' mergers.

@@

As observed by \cite{arepo}, when examining \autoref{eq:mass_transfer} one can see the relationship 
between this method and the more common (non-conservative) approach of evaluating the change in 
gas energy using the work done $(\mathbf{v} \cdot \rho \mathbf{g})$ by the gravitational force at the 
cell center. Analytically this form expresses the same core idea via the work-energy theorem, and its 
major flaw is that it evaluates the energy change at the cell center when in fact the mass transfer is 
happening at the cell edges, and the energy change is actually partially compensated by the potential 
in the adjacent zone that the mass is moving to. This is important enough that for simulations where the 
potential can change substantially over the course of the evolution, massive amounts of energy can be lost. 
This standard approach is therefore unacceptable in the case of a problem like white dwarf mergers, and 
the fix to this energy leaking -- evaluating the energy transfer at the six zone faces instead of 
the single zone center -- has a very minor cost in terms of code complexity and computational time.

@@

Another that we consider is the energy conservation of the system. Recalling \autoref{sec:gravity_hydro_coupling}, 
there are several different methods of applying the gravitational source term to the hydrodynamics equations. In \castro\ we presently 
have four options, controlled by the parameter {\tt castro.grav\_source\_type}, which we will shortern to {\tt gs} for the present 
discussion. ${\tt gs} = 1$ is the standard cell-centered source term for gravity. ${\tt gs} = 2$ is a slight variation on this scheme 
that determines the value of the energy source term after the momentum source term has been applied; the former approach uses 
the uncorrected momenta in calculating $\rho \mathbf{u} g$, which we have found to be more accurate. ${\tt gs} = 3$ is entirely 
different: after calculating the new momenta, we reset the total energy to be equal to the internal energy plus the 
kinetic energy. This has the virtue of ensuring that there is no conflict due to discretization between the momentum and 
energy equations, and also correctly ensuring that the gravitational force does not directly change the internal energy -- 
and thus the temperature -- of the fluid. However, it explicitly sacrifices total energy conservation. ${\tt gs} = 4$ is 
the new conservative method of evaluating the energy source terms at cell faces. The results for the change in energy 
after a single orbit are seen in the first column of \autoref{table:sources}. The first two versions give reasonable and 
similar levels of energy conservation. The third has total energy changes on the order of 100\%, but this itself 
does not have a severe effect on the dynamics because in this scheme the total energy variable is effectively a placeholder 
value of the kinetic energy plus internal energy, rather than being evolved directly. The last scheme is nearly two orders 
of magnitude better in energy conservation, justifying the effort in varying the scheme.

>>>>

The first attempts to model the results of the merger process came in the
1980s. \cite{nomotoiben:1985} demonstrated that off-center carbon
ignition would occur in the more massive white dwarf as it accreted
mass near the Eddington rate from the less massive white dwarf
overflowing its Roche lobe. \cite{saionomoto:1985} tracked the
evolution of the flame and found that it propagated quiescently into
the center, converting the carbon-oxygen white dwarf into an
oxygen-neon-magnesium white dwarf. This would then be followed by
collapse into a neutron star---a result with significantly different
observational properties compared to a SN Ia. This scenario, termed
accretion-induced collapse, would be avoided only if the accretion
rate were well below the Eddington rate. \cite{tutukov_yungelson:1979}
observed that this could happen if the mass loss from the secondary
was higher than the Eddington rate and thus the accreted material
formed an accretion disc, which might rain down on the primary more
slowly. The main finding was that double degenerate systems would not
obviously lead to Type Ia supernovae.

@@

A recent shift in perspective on this problem started around 2010.
\cite{pakmor:2010} used the SPH method to study the merger of 
equal-mass ($0.9\ \msolar$) carbon-oxygen white dwarfs and found 
that a hotspot was generated near the surface of the primary 
white dwarf. They argued that this region had a temperature 
and density sufficient to trigger a thermonuclear
detonation. They inserted a detonation which propagated throughout 
the system. They found that the result would observationally 
appear as a subluminous Type Ia supernova. This was the first time 
a DD simulation successfully reproduced at least some characteristics of a SN
Ia. \cite{pakmor:2011} tried a few different mass combinations and
found empirically that this would hold as long as the secondary was at
least 80\% as massive as the primary. These events, where the merger
process resulted in the detonation of the system during the merger
coalescence---avoiding the much longer time-scale evolution---were
termed ``violent'' mergers.

@@

As observed by \cite{arepo}, when examining \autoref{eq:mass_transfer} one can see the relationship 
between this method and the more common (non-conservative) approach of evaluating the change in 
gas energy using the work done $(\mathbf{v} \cdot \rho \mathbf{g})$ by the gravitational force at the 
cell center. Analytically this form expresses the same core idea via the work-energy theorem, and its 
major flaw is that it evaluates the energy change at the cell center when in fact the mass transfer is 
happening at the cell edges, and the energy change is actually partially compensated by the potential 
in the adjacent zone that the mass is moving to. This is important enough that for simulations where the 
potential can change substantially over the course of the evolution, massive amounts of energy can be lost. 
This standard approach is therefore unacceptable in the case of a problem like white dwarf mergers, and 
the fix to this energy leaking---evaluating the energy transfer at the six zone faces instead of 
the single zone center---has a very minor cost in terms of code complexity and computational time.

@@

Another that we consider is the energy conservation of the system. Recalling \autoref{sec:gravity_hydro_coupling}, 
there are several different methods of applying the gravitational source term to the hydrodynamics equations. In \castro\ we presently 
have four options, controlled by the parameter {\tt castro.grav\_source\_type}, which we will shortern to {\tt gs} for the present 
discussion. ${\tt gs} = 1$ is the standard cell-centered source term for gravity. ${\tt gs} = 2$ is a slight variation on this scheme 
that determines the value of the energy source term after the momentum source term has been applied; the former approach uses 
the uncorrected momenta in calculating $\rho \mathbf{u} g$, which we have found to be more accurate. ${\tt gs} = 3$ is entirely 
different: after calculating the new momenta, we reset the total energy to be equal to the internal energy plus the 
kinetic energy. This has the virtue of ensuring that there is no conflict due to discretization between the momentum and 
energy equations, and also correctly ensuring that the gravitational force does not directly change the internal energy---and thus the temperature---of the fluid. However, it explicitly sacrifices total energy conservation. ${\tt gs} = 4$ is 
the new conservative method of evaluating the energy source terms at cell faces. The results for the change in energy 
after a single orbit are seen in the first column of \autoref{table:sources}. The first two versions give reasonable and 
similar levels of energy conservation. The third has total energy changes on the order of 100\%, but this itself 
does not have a severe effect on the dynamics because in this scheme the total energy variable is effectively a placeholder 
value of the kinetic energy plus internal energy, rather than being evolved directly. The last scheme is nearly two orders 
of magnitude better in energy conservation, justifying the effort in varying the scheme.
